\bfseries\large Phosphatidylinositol 4,5-bisphosphate regulates cilium transition zone maturation in \textit{Drosophila melanogaster}

### 2018-02-15 latex/paper.tex
updated

In the Drosophila male germline, nascent TZs first assemble on basal bodies
during early pre-meiotic G2 phase (spermatocytes), anchor directly to the plasma membrane.
This is followed by establishment of cilium-membrane connections and a
ciliary membrane that persists during meiosis \citep{riparbelli2012assembly}
through a poorly understood mechanism
involving microtubule rearrangements \citep{gottardo2013cilium}.
Ciliogenesis begins with the assembly of TZ proteins at the tips of
the nascent basal body.
This is followed by the anchoring of the TZ to the membrane and
establishment of a compartmentalized space bounded by the ciliary membrane
and the TZ.
TZ maturation has been described in
\textit{Paramecium} \citep{aubusson2015transition}
\textit{Caenorrhabditis elegans} \citep{serwas2017centrioles} and
\textit{Drosophila melanogaster} \citep{gottardo2013cilium}
and is most readily observed by an increase in TZ length from nascent to mature.
Possibly, this is similar to the initial steps of TZ assembly and reorganization
in well-studied vertebrate cilia.

@@

\section{Discussion}
Cilia are essential for proper human development, and their disruption
can cause genetic disorders which display severity ranging from congenital
lethality to adult onset.
A large number of ciliopathies are associated with genes that encode for
proteins that function at the cilium transition zone (TZ).
Often, a single gene is linked to multiple ciliopathies with variable phenotypes
and severities.
For example, the conserved TZ protein CEP290 is linked to at least six different
cilium-associated disorders, suggesting that many of these encode for mutant proteins
that are hypomorphic in nature whose function may
be affected by modifier mutations and genetic environment.
Animal models with mild perturbations of TZ function may therefore provide
insight on intermediate phenotypes associated with human disease.
Here, we found that reduced levels of the membrane lipid \PIP{},
which is known to be enriched at the base of cilia,
affects TZ maturation, but not assembly,
leading to hyperelongated TZs that show signs of functional defects, and
may represent one such modifier for TZ function.

@@

The type I phosphatidylinositol 4-phosphate kinase (PIPKI) Skittles
suppresses TZ elongation in \sigd{} cells in a cilium-autonomous
manner.
We also found that cells mutant for \textit{sktl} contained hyperelongated
TZs similar to SigD-expressing cells.
Sktl also localizes to centrioles,
and our results suggest that Skittles act locally at cilia to
regulate TZ length.
The protein CEP135, human ortholog of the Drosophila protein
Asterless, has been show to bind PIPKI$\gamma$ in cell lines.
PIPKI$\gamma$ is linked to lethal congenital contractural syndrome
type 3 (LCCS3).
It has previously been suggested that LCCS3 might represent a ciliopathy.
PIPKI$\gamma$ is linked to LCCS3 with many features reminiscent
of ciliopathies, such as micrognathia, and a role for the LCCS1 protein
Gle3 in cilia.
In addition, all three human PIPKI isoforms are highly expressed in testes
and PIPKI$\alpha$ and PIPKI$\beta$ are important for fertility in mice
\citep{hasegawa2012phosphatidylinositol}.

Members of the exocyst complex, such as Sec10 and Sec6, have been shown
to be important for cilium formation in cultured cell lines and zebrafish
\citep{zuo2009exocyst, lobo2017exocyst, seixas2016arl13b}.
The exocyst is known to bind to \PIP{}, and we previously described that
the \textit{onr} allele of Drosophila \textit{Exo84} phenocopied
male germline defects observed in \sigd{}.
In this paper, we show that the \textit{onr} allele of the Drosophila Exo84
phenocopies the loss of cilium-membrane contacts in \sigd{},
but not TZ hyperelongation.
Proper TZ function is required for membrane tethering, as \textit{dila; cby}
mutants, which contain abnormal TZs also show cytoplasmic cilia similar to
\sigd{}.
Our results suggest that the exocyst is required after the initial stages
of TZ assembly and maturation for TZ function.

>>>>

In the Drosophila male germline, nascent TZs first assemble on basal bodies
during early G2 phase in spermatocytes and anchor directly to the plasma membrane.
This is followed by establishment of cilium-membrane connections
through a poorly understood mechanism
involving microtubule rearrangements \citep{gottardo2013cilium} and a
ciliary membrane that persists through meiosis \citep{riparbelli2012assembly}.
Ciliogenesis begins with the assembly of TZ proteins at the tips of
the nascent basal body.
This is followed by the anchoring of the TZ to the membrane and
establishment of a compartmentalized space bounded by the ciliary membrane
and the TZ.
TZ maturation has been described in
\textit{Paramecium} \citep{aubusson2015transition}
\textit{Caenorrhabditis elegans} \citep{serwas2017centrioles} and
\textit{Drosophila melanogaster} \citep{gottardo2013cilium}
and is most readily observed by an increase in TZ length from nascent to mature.
Possibly, this is similar to the initial steps of TZ assembly and reorganization
in well-studied vertebrate cilia.

@@

\section{Discussion}
Cilia are essential for proper human development, and their disruption
can cause genetic disorders which display severity ranging from congenital
lethality to adult onset.
A large number of ciliopathies are associated with genes that encode for
proteins that function at the cilium transition zone (TZ).
Often, a single gene is linked to multiple ciliopathies with distinct phenotypes
and severities.
For example, \textit{CEP290} is linked to at least six different
cilium-associated disorders.
It is likely that these mutations do not completely abolish CEP290
function, and may be affected by genetic modifiers in an affected individual
\citep{zaghloul2010functional}.
Animal models with partial loss of TZ function may therefore provide
insight on intermediate phenotypes associated with human disease.
Here, we found that reduced levels of the membrane lipid \PIP{},
which is known to be enriched at the base of cilia,
affects TZ maturation, but not assembly,
leading to hyperelongated TZs that show signs of functional defects, and
may represent one such modifier for TZ function.

@@

The type I phosphatidylinositol 4-phosphate kinase (PIPKI) Skittles
suppresses TZ elongation in \sigd{} cells in a cilium-autonomous
manner.
We also found that cells mutant for \textit{sktl} contained hyperelongated
TZs similar to SigD-expressing cells.
Sktl also localizes to centrioles,
and our results suggest that Skittles acts locally at cilia to
regulate TZ length.
PIPKI$\gamma$ is linked to lethal congenital contractural syndrome
type 3 (LCCS3).
It has previously been suggested that LCCS3 might represent a ciliopathy.
PIPKI$\gamma$ is linked to LCCS3 with features reminiscent
of ciliopathies such as micrognathia, and a role for the LCCS1 protein
Gle3 in cilia.
In addition, all three human PIPKI isoforms are highly expressed in testes
and PIPKI$\alpha$ and PIPKI$\beta$ are important for fertility in mice
\citep{hasegawa2012phosphatidylinositol}.

Members of the exocyst complex, such as Sec10 and Sec6, have been shown
to be important for cilium formation in cultured cell lines and zebrafish
\citep{zuo2009exocyst, lobo2017exocyst, seixas2016arl13b}.
The exocyst is known to bind to \PIP{}
\citep{he2007exo70, zhang2008membrane}, and we previously described that
the \textit{onr} allele of Drosophila \textit{Exo84} phenocopied
male germline defects observed in \sigd{} \citep{fabian2010phosphatidylinositol}.
In this paper, we show that the \textit{onr} allele of the Drosophila Exo84
phenocopies the loss of cilium-membrane contacts in \sigd{},
but not TZ hyperelongation.
Proper TZ function is required for membrane tethering, as \textit{dila; cby}
mutants, which contain abnormal TZs also show cytoplasmic cilia similar to
\sigd{} \citep{vieillard2016transition}.
Our results suggest that the exocyst is required after the initial stages
of TZ assembly and maturation for TZ function.
